Distributional Trees for Circular Data

### 2019-04-09 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
further revisions on extended abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2521 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: e1ab82ca1c016aa6e46d2c39bf7127efb1bfb047

Circular data can be found in a variety of applications and subject areas, 
e.g., hourly crime rate in the social-economics, 
animal movement direction or gene-structure in biology, 
and wind direction as one of the most important weather variables in meteorology.
Circular regression models were first introduced by Fisher and Lee (1992) and 
further extended by Jammalamadaka and Sengupta (2001) and 
Mulder and Klugkist (2017) among others.
While most of these approaches are built on additive regression models, 
we propose an adaption/extension of regression trees to circular data by employing 
distributional trees.

@@

\subsection{Ensembles/Random Forests}
A natural extension are ensembles (or forests) of such circular trees
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by Breiman~et.~al~(1984) average the
predictions of an ensemble of trees, each built on a subsample 
or bootstrap of the original data. 
A different strategy of combining trees is employed in distributional forests 
(Schlosser~et.~al, 2019) which can also be seen as adaptive local likelihood models. 
For a possibly new observation $x$ weights $w_i(x)$ are obtained by counting the 
number of trees in which it is assigned to the same terminal node as a learning
observation $y_i, i \in \{1,\ldots,n\}$.
The parameters $\mu$ and $\kappa$ are then estimated specifically for the (new) 
observation $x$ by applying the weighted maximum likelihood estimator
$$
\operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
$$
In that way predictions are mainly based on similar observations rather than the 
complete learning data.

\subsection{Splits in Circular Variables}
Another planned extension of the introduced circular distributional trees 
for obtaining more parsimonious and more stable trees is to
consider splitting circular covariates into two flexible circle segments by searching two split 
points simultaneously.
In general, the search for the best separation of the data space into two intervals is based on the optimization problem
$$
\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
$$
where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters for the corresponding subsets of the response as defined by the intervals, here generally denoted by \textit{``left''} and \textit{``right''}.
For a linear split variable these intervals are defined by a single split point.
By treating a circular variable such as wind direction linearly
this single split point $\nu$ devides the considered space~$[0,2\pi)$ into
$\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$ enforcing a potentially unnatural 
separation at zero. 
Moreover, further sequential splits in the circular variable are often required to obtain 
data driven subintervals of~$[0,2\pi)$ (as seen for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
This can be avoided by searching for two split points $\nu$ and $\tau$ within the maximization
process such that $\mathit{left}=[\nu,\tau]$ and $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$
where both interval borders are determined purely by the data.

>>>>

Circular data can be found in a variety of applications and subject areas, 
e.g., hourly crime rate in the social-economics, 
animal movement direction or gene-structure in biology, 
and wind direction as one of the most important weather variables in meteorology.
Circular regression models were first introduced by Fisher and Lee (1992) and 
further extended by Jammalamadaka and Sengupta (2001) and 
Mulder and Klugkist (2017) among others.
While most of the already existing approaches are built on additive regression models, 
we propose an adaption/extension of regression trees to circular data by employing 
distributional trees.

@@

\subsection{Ensembles/Random forests}
A natural extension are ensembles (or forests) of such circular trees
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by Breiman~et.~al~(1984) average the
predictions of an ensemble of trees, each built on a subsample 
or bootstrap of the original data. 
A different strategy of combining trees is employed in distributional forests 
(Schlosser~et.~al, 2019) which can also be interpreted as adaptive local likelihood models. 
For a possibly new observation $x$ averaged nearest neighbor weights $w_i(x)$ are obtained 
by counting the number of trees in which it is assigned to the same terminal node as a 
learning observation $y_i, i \in \{1,\ldots,n\}$ divided by the total number of trees.
The parameters $\mu$ and $\kappa$ are then estimated specifically for the (new) 
observation $x$ by applying the weighted maximum likelihood estimator
\begin{equation}
\operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
\end{equation}
In that way predictions are mainly based on similar observations rather than the 
complete learning data.

\subsection{Splits in circular variables}
In order to obtain more parsimonious and more stable trees another possible extension of the 
introduced circular distributional trees is to split circular covariates into two 
flexible circle segments by searching two split points simultaneously.
In general, the search for the best separation of the data space into two intervals is based 
on the optimization problem
\begin{equation}
\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
\end{equation}
where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters
of the von Mises distribution for the corresponding subsets of the response as defined by the
intervals, here generally denoted by \textit{``left''} and \textit{``right''}.
For a linear split variable these intervals are defined by a single split point.
By treating a circular variable such as wind direction linearly
this single split point $\nu$ divides the considered space~$[0,2\pi)$ into
$\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$ enforcing a potentially unnatural 
separation at zero. 
Moreover, further sequential splits in the circular variable are often required to obtain 
data driven subintervals of~$[0,2\pi)$ (as seen for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
This can be avoided by searching for two split points $\nu$ and $\tau$ within the maximization
process such that $\mathit{left}=[\nu,\tau]$ and $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$
where both interval borders are determined purely by the data.
